### 2010-03-30 papers/icfp-2010-desc/paper_idesc.tex
Desc@ICFP: fix typos in Section 4, 5.4, 5.5, 6, and 7

However, this new class of objects cannot be defined by mere
induction. In the case of vectors, for instance, we have to define the
whole \emph{family} of vector in one go: vectors of all sizes need to
be defined at the same time. Where simply-typed languages have
inductive types, in dependently typed programming the basic grammar of
data-types should be that of inductive families of types. To this end, we
rely on \emph{indexing}. We are going to transform the $\Desc$ universe into
$\IDesc{\!}$, the universe of indexed descriptions. We express inductive families
by the $\V{I} \To \IDesc{\V{I}}$ type.  \note{pwm: Vec might not be the best
motivating example here, since it doesn't appear in the paper (more on that
later). Come to think of it, this is entirely too vague without spelling out
a concrete example.}

@@

Before moving on to indexed descriptions, we have to carry out some
maintenance work on descriptions. We presented $\Desc$ as the grammar
of inductive types. Hence, the codes closely follow this grammar. In
the following, we adopt an alternative presentation. With
$\DSigma{\!}{\!}$, we are actually \emph{quoting} a standard
type-former, namely

@@

\[\stk{
\data \IDesc{(\Bhab{\V{I}}{\Set})} : \Set \where \:\ldots \\
\\
\idescop{\_}{}{} : _{\PI{\V{I}}{\Set}} \IDesc{\V{I}} \To (\V{I} \To \Set) \To \Set    \\
\ldots
}\]

Note that $\V{I} \To \IDesc{\V{I}}$ is isomorphic, up to interpretation, to an
endofunctor on $\Set^{\V{I}}$. Indeed, $\V{I} \To \IDesc{\V{I}}$ is interpreted as
$\V{I} \To \Set^{\V{I}} \To \Set$, which is isomorphic to $\Set^{\V{I}}
\To \Set^{\V{I}}$. Fix-point of inductive families is defined over
these indexed functors, hence computing a fix-point of the entire
\emph{family} of functors:

\[\stkl{
\Rule{\Gamma \vdash \Bhab{\V{I}}{\Set} \qquad
      \Gamma \vdash \Bhab{\V{R}}{\V{I} \To \IDesc{\V{I}}}}
     {\Gamma \vdash \Bhab{\IMu{\V{I}}{\V{R}}{\!}}{\V{I}\To\Set}} \qquad
\\:\
\Rule{\begin{array}{l@{\qquad}l}
          \Gamma \vdash \Bhab{\V{I}}{\Set} &
          \Gamma \vdash \Bhab{\V{R}}{\V{I} \To \IDesc{\V{I}}} \\
          \Gamma \vdash \Bhab{\V{i}}{\V{I}} &
          \Gamma \vdash \Bhab{\V{x}}{\idescop{\V{R}\:\V{i}}{\V{I}}{(\IMu{\V{I}}{\V{R}}{\!})}}
      \end{array}}
     {\Gamma \vdash \Bhab{\Con{\V{x}}}{\IMu{\V{I}}{\V{R}}{\V{i}}}}
}\]

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \IDesc{(\Bhab{\V{I}}{\Set})} : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DVar{\!}         & \V{I} \To \IDesc{\V{I}}                                       \\
    \DConst{\!}       & \Set \To \IDesc{\V{I}}                                        \\
    \DProd{\!}{\!}    & \PI{\V{\V{D}}, \V{\V{D'}}}{\IDesc{\V{I}}} \IDesc{\V{I}}       \\
    \DSigma{\!}{\!}   & \PI{\V{S}}{\Set} \PIS{\V{S} \To \IDesc{\V{I}}} \IDesc{\V{I}}  \\
    \DPi{\!}{\!}      & \PI{\V{S}}{\Set} \PIS{\V{S} \To \IDesc{\V{I}}} \IDesc{\V{I}} 
\end{array}
}
\vspace{0.2in}
\\
\stk{
\idescop{\_\:}{}{} :_{\PI{\V{I}}{\Set}} \IDesc{\V{I}} \To (\V{\V{I}} \To \Set) \To \Set                  \\
\begin{array}{@{}l@{\:=\:\:}ll}
\idescop{\DVar{\V{i}}}{\V{I}}{\V{P}}      &  \V{P}~\V{i}                                                 \\
\idescop{\DConst{\V{X}}}{\V{I}}{\V{P}}    &  \V{X}                                                   \\
\idescop{\DProd{\V{D}}{\V{D'}}}{\V{I}}{\V{P}} &  \TIMES{\idescop{\V{D}}{\V{I}}{\V{P}}}{\idescop{\V{D'}}{\V{I}}{\V{P}}}       \\
\idescop{\DSigma{\V{S}}{\V{D}}}{\V{I}}{\V{P}} &  \SIGMA{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{P}}                    \\
\idescop{\DPi{\V{S}}{\V{D}}}{\V{I}}{\V{P}}    &  \PI{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{P}}            
\end{array}
}
\end{array}
}\]

@@

However, one could, quite rightfully, be worried by the introduction
of equalities in the picture. In a programming context, equalities
haunt us, begging for proofs. Hence, whereas the developer typed a
constructor-constrained data-type, the elimination form will involve
equalities and their proofs. As much as possible, one would rather
avoid suffering from this burden. However, the equalities introduced
here are rather special: they are \emph{equations} constraining
variables to certain terms. \citet{goguen:pattern-matching} have
shown, in the context of dependent pattern matching, how to tame these
equations.

@@

Further, we can actually remove these equations altogether. Let us
look back at $\D{Vec}$. We note that the equations are introduced
because we are \emph{storing} the index of the inductive family.
However, \emph{inductive families need not store their
  indices}~\cite{brady:index-inductive-families}.  By examining the
incoming index, we can apply the \emph{forcing} and \emph{de-tagging}
optimisations to our initial definition of $\D{Vec}$ gives the
following, equivalent definition:

@@

\[
\stk{
\data \Fin{\!} : \PI{\V{n}}{\Nat} \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \FinZero      & _{\PI{\V{n}}{\Nat}}\Fin{(\NatSuc{\V{n}})}   \\
    \FinSuc{\!}   & _{\PI{\V{n}}{\Nat}}\Fin{\V{n}} \To \Fin{(\NatSuc{\V{n}})}
\end{array}
\vspace{0.1in}
\\
\FinD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}llll}
\FinD\: \NatZero     & = & \multicolumn{2}{l}{\DSigma{(\EnumT{[]})}{\Void} } \\
\FinD\: (\NatSuc{\V{n}}) & = & \DSigma{\!}{\!}\!\!\!\!\! & (\EnumT{[ \FinZero , \FinSuc{\!} ]}) \\
                     &   &             & \bigRedBracket{
                                         \begin{array}{l}
                                         \DConst{\Unit} \\
                                         \DVar{\V{n}}
                                         \end{array}
                                         }
\end{array}
}\]

@@

For convenience, we divide a tagged indexed description in two parts:
first, the constructors that do not depend on the index; then, the
constructors that do. The non-dependent part mirrors the definition
for non-indexed descriptions: we are provided a finite choice of
constructors. The index-depend part simply indexes the choice of
constructors by $\V{I}$. Hence, by inspecting the index, it is
possible to enable or disable the constructors.

@@

%% Index mapper (terminology?)
\newcommand{\Val}[1]{\F{Val}~#1}
\newcommand{\Var}[2]{\F{Var}_{#1}~#2}

%% Hutton expressions
\newcommand{\HExprD}{\C{ExprD}}
\newcommand{\HExprAD}{\C{ExprAD}}
\newcommand{\HExprID}{\C{ExprID}}
\newcommand{\HExprVarD}[1]{\C{ExprD}_{\F{Var},#1}}
\newcommand{\HExprFreeD}{\C{ExprD}^{\C{Free}}}
\newcommand{\HExprAFreeD}{\C{ExprAD}^{\C{Free}}}

So far, the examples we have seen lives in $\IDesc{\Unit}$, hence are
not using any indexing. We are going to define a syntax for a small
typed language. We consider two types, natural numbers and booleans:

@@

The function $\Val{}$, used in the definition of $\Eval{}$, simply
maps an object type $ty$ to the corresponding type in the host
language. Hence, the argument of $\Eval{}$ are ensured to be of the
expected type. We assume $\Nat$ and $\Bool$ represent natural numbers
and booleans in the host language. Further, we also assume the
existence of an addition and comparison operator in $\Nat$,
respectively named $\F{plusHost}$ and $\F{leHost}$. We define $\Val{}$
as follows:

@@

In our universe of descriptions, this data-type is represented by a
tagged indexed description. We use the index to carry the type: the
resulting description is indexed by $\Ty$. We observe that some
constructors are ``polymorphic'', namely $\Econd{}{}{}$ and $\Eval{}$. On the
other hand, the $\Eplus$ and $\Ele$ constructors are
index-dependent. $\Eplus$ is defined if and only if the result type --
the index -- is $\Enat$, whereas $\Ele$ is defined if and only if the
index is $\Ebool$. The actual code precisely follows this intuition,
as shown in Figure~\ref{fig:hexpr-full}. For brevity, we use an
informal $\case{\ldots}{\ldots}$ notation, simulating a definition by
pattern-matching. Formally, this corresponds to a call to the
$\switch{\!}{\!}{\!}{\!}$ eliminator.

@@

\[\stk{
\stk{
\HExprD : \TagIDesc{\Ty} \\
\HExprD \mapsto ( \HExprAD , \HExprID ) \\
} \\
\\
\stk{
\HExprAD : \ATagIDesc{\Ty} \\
\HExprAD \mapsto \bigRedBracket{
                 \begin{array}{l}
                   {[\Eval{}, \Econd{\!}{\!}{\!}\!\,]} \red{,} \\
                   \LAM{\V{ty}}
                   \bigRedBracket{
                   \begin{array}{l}
                   \DConst{(\Val{\V{ty}})} \\
                   \DProd{\DVar{\Ebool}}{\DProd{\DVar{\V{ty}}}{\DVar{\V{ty}}}}
                   \end{array}
                   }
                 \end{array}
                 }
\\
\\
} 
\\
\stk{
\HExprID : \ITagIDesc{\Ty} \\
\HExprID \mapsto \bigRedBracket{
                 \begin{array}{l}
                   \LAM{\V{ty}} \case{\V{ty}}{\Enat \To {[\Eplus{\!}{\!}\!\,]} \\ \Ebool \To {[\Ele{\!}{\!}\!\,]}} \red{,} \\
                   \LAM{\_} \DProd{\DVar{\Enat}}{\DVar{\Enat}} 
                   \end{array}
                   }
}
}\]

@@

Where we overload the notation $\toIDesc{\V{TID}}$ to denote the
indexed description computed from the tagged indexed description
$\V{TID}$.

The type of $\F{eval}$ is strikingly similar to a
catamorphism. Indeed, implementing a single step of evaluation -- the
algebra -- is sufficient, as $\F{cataI}$ \note{We have not defined
  \F{cataI}!} gives, for free, the full evaluator. The implementation
is as follows:

\[\stk{
\evalOne : \PI{\V{ty}}{\Ty} \idescop{\toIDesc{\HExprD}\:\V{ty}}{\Ty}{\Val{\!}}
\To {\Val{\V{ty}}} \\
\begin{array}{@{}l@{\:=\:\:}l}
\evalOne\: \_\: \pair{\Eval{}}{\V{x}}{}                                         & \V{x} \\
\evalOne\: \_\: \pair{\Econd{}{}{}}{\pair{\BoolTrue}{\pair{\V{x}}{\_}{}}{}}{}   & \V{x} \\
\evalOne\: \_\: \pair{\Econd{}{}{}}{\pair{\BoolFalse}{\pair{\_}{\V{y}}{}}{}}{}  & \V{x} \\
\evalOne\: \Enat\: \pair{\Eplus{}{}}{\pair{\V{x}}{\V{y}}{}}{}                       & \F{plusHost}\: \V{x}\: \V{y} \\
\evalOne\: \Ebool\: \pair{\Ele{}{}}{\pair{\V{x}}{\V{y}}{}}{}                        & \F{leHost}\: \V{x}\: \V{y} 
\end{array} \\
\\
\evalH : \PI{\V{ty}}{\Ty} 
           \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}} \To
           \Val{\V{ty}} \\
\evalH\: \V{ty}\: \V{term} = \F{cataI}\: \Ty\: 
                                 \toIDesc{\HExprD}\: 
                                 \Val{}\: 
                                 \evalOne\: 
                                 \V{ty}\: 
                                 \V{term}
}\]
\note{pwm: $\evalOne$ rather illustrates my point about the 2 types of
  IFunc floating around, $\idescop{\_}{}{}$ is defined using the 1st
  notion, and used it here as if it were the 2nd. pierre: in this
  case, it was a typo. But you're right, there is two notions and we
  probably want to clarify where these things live.}

@@

However, so far, we are only able to define and manipulate
\emph{closed} terms. By abstracting over $\Val{}$, it is possible to
build and manipulate \emph{open} terms, that is, terms with symbolic
variables. On the model of $\Val{}$, we define $\Var{}{}$:

@@

Consequently, replacing $\Val{\V{ty}}$ by
$(\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}})$ in Figure~\ref{fig:hexpr-full} turns
the language of closed term into a language of opened terms with
constants. For readability, we will abbreviate $\LAM{\V{ty}}
\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}}$ into $\SUM{\Val{\!}}{\Var{\V{dom}}{\!}}$. \note{Is
  it polite and comprehensible to ask that?} This defines a new
indexed description, called $\HExprVarD{\V{dom}}$.

@@

\[
\begin{array}{@{}ll}
\substH  : & \PITEL{\V{dom}}{\EnumU} \\
           & \PITEL{\V{\gammaNat}}{\GammaSpi{\V{dom}}{\Enat}} 
             \PITEL{\V{\gammaBool}}{\GammaSpi{\V{dom}}{\Ebool}} \\
           & \begin{array}{@{}ll}
             \PI{\V{\sigma}}{& \PITEL{\V{dom}}{\EnumU} \\
                             & \PITEL{\V{\gammaNat}}
                                     {\GammaSpi{\V{dom}}{\Enat}} \\
                             & \PITEL{\V{\gammaBool}}
                                     {\GammaSpi{\V{dom}}{\Ebool}} \\
                             & \PI{\V{ty}}{\Ty} (\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}}) \To \\
                             & \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}}}
             \end{array}\\
          & \PI{\V{ty}}{\Ty}
            \IMu{\Ty}{\toIDesc{\HExprVarD{\V{dom}}}}{\V{ty}} \To \\
          & \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}}
\end{array}
\]

@@

Abstracting away the book-keeping introduced by the context, this
definition looks familiar. Indeed, it is extremely similar to a
monadic \bind. This is not surprising as we are defining a first-order
syntax with variables: our data-type enjoys more structure than what
we are given. In particular, this is reminiscent of a free monad,
where $\Eval{}$ is the \return\ introducing variables. The substitution
$\V{\sigma}$ is implemented from $\discharge$, by picking the domain and
context corresponding to the type:

@@

Instead of implementing $\substH$ in this special case, we shall
implement it in a generic setting, and express our example in that
setting. However, before entering the realm of generic structures, let
us further explore the universe of indexed descriptions.

\subsection{Constrained constructors}

\begin{wstructure}
!!! Need Help !!!
<- Assuming a suitable notion of definitional equality for I
    ???
\end{wstructure}

In the following, we ask $I$ to come equipped with a notion of
definitional equality. Hence, we can legibly equates elements of $I$.

\note{ I will need some tutorial on this ad-hoc notion of definitional
  equality. Or what Conor's meant with that. }

\begin{wstructure}
<- Fin [figure]
    <- Presentation with equality constraints
    <- Example of constrained constructor
        -> Constraints are translated into equalities
    ???
\end{wstructure}

Another common dependent data-type is $\Fin{\!}$, the finite sets of
numbers. $\Fin{\!}$ is a typical example of an inductive family using
\emph{constrained constructors}. The constructors $\FinZero$ and
$\FinSuc{\!}$ are only defined for an index distinct from $\NatZero$:

\[
\stk{
\data \Fin{\!} : \PI{\V{n}}{\Nat} \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \FinZero_{\V{n}}      & \Fin{(\NatSuc{\V{n}})}   \\
    \FinSuc{\!}_{\V{n}}   & \Fin{\V{n}} \To \Fin{(\NatSuc{\V{n}})}
\end{array}
}
\]

\[\stk{
\FinD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}ll}
\FinD\: \V{n} = \DSigma{\!}{\!} \!\!\!\!& (\EnumT{[ \FinZero , \FinSuc{\!} ]}) \\
                        & \bigRedBracket{
                          \begin{array}{l}
                            \DSigma{\Nat}{\LAM{\V{m}} \DConst{(\V{n} \PropEq \NatSuc{\V{m}})}} \\
                            \DSigma{\Nat}{\LAM{\V{m}} \DProd{\DVar{\V{m}}}{\DConst{(\V{n} \PropEq \NatSuc{\V{m}})}}}
                          \end{array}
                          }
\end{array}
}\]

Intuitively, the constraint is captured as follows. First, we store an
element $\V{m}$ of $\Nat$, thanks to a $\DSigma{\!}{\!}$. However, the
constraint stipulates that $\V{m}$ cannot be \emph{any} natural
numbers: it must be ``the index minus one''. This translates into the
constraint $\V{n} \PropEq \NatSuc{\V{m}}$.

\note{ Anything else? }

\begin{wstructure}
<- GADT-style data-type definition
    <- We directly support it
        <- Constraints turned into equalities
            <- Fin example
    <- Again, smooth transition from "today" data-types to advanced ones
        <- Same grammar
        /> More power
    /> Equations in data-types are not a blessing
        <- Complexify induction principle
        <- Lose match between data-type definition and the elimination form
\end{wstructure}

The $\Fin{\!}$ example is also a well-known example of Generalised
Abstract Data-Type (GADT) in Haskell. It comes as no surprise that our
universe of indexed description supports this restricted class of
inductive families. As we have seen with $\Fin{\!}$, constrained
constructors are straightforwardly translated into equations.

Further, we can actually remove these equations altogether. Let us
look back at $\Fin{\!}$. We note that the equations are introduced because
we are \emph{storing} the index of the inductive family, through
$\V{m}$. Hence the constraints. However, \emph{inductive families need not
  store their indices}~\cite{brady:index-inductive-families}. Applying
the \emph{forcing} optimisation to our initial definition of $\Fin{\!}$
gives the following, equivalent definition:

\[\stk{
\FinD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}llll}
\FinD\: \NatZero     & = & \multicolumn{2}{l}{\DSigma{(\EnumT{[]})}{\Void} } \\
\FinD\: (\NatSuc{\V{n}}) & = & \DSigma{\!}{\!}\!\!\!\!\! & (\EnumT{[ \FinZero , \FinSuc{\!} ]}) \\
                     &   &             & \bigRedBracket{
                                         \begin{array}{l}
                                         \DConst{\Unit} \\
                                         \DVar{\V{n}}
                                         \end{array}
                                         }
\end{array}
}\]

Equations have simply disappeared. Note that it is straightforward to
turn this definition into a tagged form. We should precise that
forcing a description is not guaranteed to remove all constraints. It
is subject to future work to see if constraints can be entirely
eradicated, or presented more conveniently to the developer. Finally,
it is worth mentioning that the forcing optimisation is a
\emph{source-to-source} transformation of the description.

@@

In this section, we develop more structure for our indexed
data-types. In Section~\ref{sec:desc-free-monad}, we have constructed
a free monad operation on simple descriptions. Building on this
experience, we are going to present its equivalent in the indexed
world. The process the same. Namely, given an indexed functor, we
derive the indexed functor coding its free monad:
\note{pwm: Whoa there. Maybe we should say something about IMonads in
general before we get to this point?}

>>>>

However, this new class of objects cannot be defined by mere
induction. In the case of vectors, for instance, we have to define the
whole \emph{family} of vector in one go: vectors of all size needs to
be defined at the same time. Where simply-typed languages have
inductive types, in dependently typed programming we the basic grammar of
data-types should be that of inductive families of types. To this end, we
rely on \emph{indexing}. We are going to transform the $\Desc$ universe into
$\IDesc{\!}$, the universe of indexed descriptions. We express inductive families
by the $\V{I} \To \IDesc{\V{I}}$ type.  \note{pwm: Vec might not be the best
motivating example here, since it doesn't appear in the paper (more on that
later). Come to think of it, this is entirely too vague without spelling out
a concrete example.}

@@

Before moving on indexed descriptions, we have to carry out some
maintenance work on descriptions. We presented $\Desc$ as the grammar
of inductive types. Hence, the codes closely follow this grammar. In
the following, we adopt an alternative presentation. With
$\DSigma{\!}{\!}$, we are actually \emph{quoting} a standard
type-former, namely

@@

\[\stk{
\data \IDesc{\!} (\Bhab{\V{I}}{\Set}) : \Set \where \\
\;\;\ldots \\
\\
\idescop{\_\:}{}{} : \PI{\V{I}}{\Set} \IDesc{\V{I}} \To (\V{I} \To \Set) \To \Set    \\
\ldots
}\]

Note that $I \To \IDesc{\V{I}}$ is isomorphic, up to interpretation, to an
endofunctor on $\Set^{\V{I}}$. Indeed, $\V{I} \To \IDesc{\V{I}}$ is interpreted as
$\V{I} \To (\V{I} \To \Set) \To \Set$, which is isomorphic to $(\V{I} \To \Set)
\To (\V{I} \To \Set)$. Fix-point of inductive families is defined over
these indexed functors, hence computing a fix-point of the entire
\emph{family} of functors:

\[\stkl{
\Rule{\stk{ 
          \Gamma \vdash \Bhab{\V{I}}{\Set} \\
          \Gamma \vdash \Bhab{\V{R}}{\V{I} \To \IDesc{\V{I}}} \\
          \Gamma \vdash \Bhab{\V{i}}{\V{I}}}}
     {\Gamma \vdash \Bhab{\IMu{\V{I}}{\V{R}}{\V{i}}}{\Set}} \qquad
\Rule{\stk{
          \Gamma \vdash \Bhab{\V{I}}{\Set} \\
          \Gamma \vdash \Bhab{\V{R}}{\V{I} \To \IDesc{\V{I}}} \\
          \Gamma \vdash \Bhab{\V{i}}{\V{I}} \\
          \Gamma \vdash \Bhab{\V{x}}{\idescop{\V{R}\: \V{i}}{\V{I}}{(\IMu{\V{I}}{\V{R}}{})}}}}
     {\Gamma \vdash \Bhab{\Con{\V{x}}}{\IMu{\V{I}}{\V{R}}{\V{i}}}}
}\]

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \IDesc{\!} (\Bhab{\V{I}}{\Set}) : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DVar{\!}         & \V{I} \To \IDesc{\V{I}}                                       \\
    \DConst{\!}       & \Set \To \IDesc{\V{I}}                                        \\
    \DProd{\!}{\!}    & \PI{\V{\V{D}}, \V{\V{D'}}}{\IDesc{\V{I}}} \IDesc{\V{I}}       \\
    \DSigma{\!}{\!}   & \PI{\V{S}}{\Set} \PIS{\V{S} \To \IDesc{\V{I}}} \IDesc{\V{I}}  \\
    \DPi{\!}{\!}      & \PI{\V{S}}{\Set} \PIS{\V{S} \To \IDesc{\V{I}}} \IDesc{\V{I}} 
\end{array}
}
\vspace{0.2in}
\\
\stk{
\idescop{\_\:}{}{} :_{\PI{\V{I}}{\Set}} \IDesc{\V{I}} \To (\V{\V{I}} \To \Set) \To \Set                  \\
\begin{array}{@{}l@{\:=\:\:}ll}
\idescop{\DVar{\V{i}}}{\V{I}}{\V{P}}      &  \V{P}~\V{i}                                                 \\
\idescop{\DConst{\V{X}}}{\V{I}}{\V{P}}    &  \V{X}                                                   \\
\idescop{\DProd{\V{D}}{\V{D'}}}{\V{I}}{\V{P}} &  \TIMES{\idescop{\V{D}}{\V{I}}{\V{P}}}{\idescop{\V{D'}}{\V{I}}{\V{P}}}       \\
\idescop{\DSigma{\V{S}}{\V{D}}}{\V{I}}{\V{P}} &  \SIGMA{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{P}}                    \\
\idescop{\DPi{\V{S}}{\V{D}}}{\V{I}}{\V{P}}    &  \PI{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{P}}            
\end{array}
}
\end{array}
}\]

@@

@@

Further, we can actually remove these equations altogether. Let us look back
at $\D{Vec}$. We note that the equations are introduced because we are
\emph{storing} the index of the inductive family.  However, \emph{inductive
families need not store their indices}~\cite{brady:index-inductive-families}.
By examining the in coming index, we can apply the \emph{forcing} and
\emph{de-tagging} optimisations to our initial definition of $\D{Vec}$ gives
the following, equivalent definition:

@@

\[
\stk{
\data \Fin{\!} : \PI{\V{n}}{\Nat} \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \FinZero_{\V{n}}      & \Fin{(\NatSuc{\V{n}})}   \\
    \FinSuc{\!}_{\V{n}}   & \Fin{\V{n}} \To \Fin{(\NatSuc{\V{n}})}
\end{array}
\vspace{0.1in}
\\
\FinD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}llll}
\FinD\: \NatZero     & = & \multicolumn{2}{l}{\DSigma{(\EnumT{[]})}{\Void} } \\
\FinD\: (\NatSuc{\V{n}}) & = & \DSigma{\!}{\!}\!\!\!\!\! & (\EnumT{[ \FinZero , \FinSuc{\!} ]}) \\
                     &   &             & \bigRedBracket{
                                         \begin{array}{l}
                                         \DConst{\Unit} \\
                                         \DVar{\V{n}}
                                         \end{array}
                                         }
\end{array}
}\]

@@

For convenience, we divide a tagged indexed description in two parts:
first, the constructors that do not depend on the index; then, the
constructors that do. The non-dependent part mirrors the definition
for non-indexed descriptions: we are provided a finite choice of
constructors. The index-depend part simply indexes the choice of
constructors by $\V{I}$. Hence, by inspecting the index, it is possible to
e

@@

%% Index mapper (terminology?)
\newcommand{\Val}[1]{\D{Val}~#1}
\newcommand{\Var}[2]{\D{Var}_{#1}~#2}

%% Hutton expressions
\newcommand{\HExprD}{\C{ExprD}}
\newcommand{\HExprAD}{\C{ExprAD}}
\newcommand{\HExprID}{\C{ExprID}}
\newcommand{\HExprVarD}[1]{\C{ExprD}_{\C{Var},#1}}
\newcommand{\HExprFreeD}{\C{ExprD}^{\C{Free}}}
\newcommand{\HExprAFreeD}{\C{ExprAlwaysD}^{\C{Free}}}

We are going to define a syntax for a small
typed language. We consider two types, natural numbers and booleans:

@@

The function $\Val{}$, used in the definition of $\Eval{}$, simply
maps an object type $ty$ to the corresponding type in the host
language. Hence, the argument of $\Eval{}$ are ensured to be of the
expected type. We assume $\Nat$ and $\Bool$ represent natural numbers
and booleans in the host language. Further, we also assume the
existence of an addition and comparison operator in $\Nat$,
respectively named $\F{plusHost}$ and $\F{leHost}$. We define $\Val{}$
as follow:

@@

In our universe of descriptions, this data-type is represented by a
tagged indexed description. We use the index to carry the type: the
resulting description is indexed by $\Ty$. We observe that some
constructors are ``polymorph'', namely $\Econd{}{}{}$ and $\Eval{}$. On the
other hand, the $\Eplus$ and $\Ele$ constructors are
index-dependent. $\Eplus$ is defined if and only if the result type --
the index -- is $\Enat$, whereas $\Ele$ is defined if and only if the
index is $\Ebool$. The actual code precisely follows this intuition,
as shown in Figure~\ref{fig:hexpr-full}. For brevity, we use an
informal $\case{\ldots}{\ldots}$ notation, simulating a definition by
pattern-matching. Formally, this corresponds to a call to the
$\switch{\!}{\!}{\!}{\!}$ eliminator.

@@

\[\stk{
\stk{
\HExprD : \TagIDesc{\Ty} \\
\HExprD \mapsto ( \HExprAD , \HExprID ) \\
} \\
\\
\stk{
\HExprAD : \ATagIDesc{\Ty} \\
\HExprAD \mapsto \bigRedBracket{
                 \begin{array}{l}
                   \EnumT{[\Eval{}, \Econd{\!}{\!}{\!}\!\,]} \red{,} \\
                   \LAM{\V{ty}}
                   \bigRedBracket{
                   \begin{array}{l}
                   \DConst{(\Val{\V{ty}})} \\
                   \DProd{\DVar{\Ebool}}{\DProd{\DVar{\V{ty}}}{\DVar{\V{ty}}}}
                   \end{array}
                   }
                 \end{array}
                 }
\\
\\
} 
\\
\stk{
\HExprID : \ITagIDesc{\Ty} \\
\HExprID \mapsto \bigRedBracket{
                 \begin{array}{l}
                   \LAM{\V{ty}} \case{\V{ty}}{\Enat \To \EnumT{[\Eplus{\!}{\!}\!\,]} \\ \Ebool \To \EnumT{[\Ele{\!}{\!}\!\,]}} \red{,} \\
                   \LAM{\_} \DProd{\DVar{\Enat}}{\DVar{\Enat}} 
                   \end{array}
                   }
}
}\]

@@

The type of $\F{eval}$ is strikingly similar to a
catamorphism. Indeed, implementing a single step of evaluation -- the
algebra -- is sufficient, as $\F{cataI}$ \note{We have not defined
  \F{cataI}!} gives, for free, the full evaluator. The implementation
is as follow:

\[\stk{
\evalOne : \PI{\V{ty}}{\Ty} \idescop{\toIDesc{\HExprD}\: \V{ty}}{\Ty}{\V{ty}}
\To {\Val{\V{ty}}} \\
\begin{array}{@{}l@{\:=\:\:}l}
\evalOne\: \_\: \pair{\Eval{}}{\V{x}}{}                                         & \V{x} \\
\evalOne\: \_\: \pair{\Econd{}{}{}}{\pair{\BoolTrue}{\pair{\V{x}}{\_}{}}{}}{}   & \V{x} \\
\evalOne\: \_\: \pair{\Econd{}{}{}}{\pair{\BoolFalse}{\pair{\_}{\V{y}}{}}{}}{}  & \V{x} \\
\evalOne\: \Enat\: \pair{\Eplus{}{}}{\pair{\V{x}}{\V{y}}{}}{}                       & \F{plusHost}\: \V{x}\: \V{y} \\
\evalOne\: \Ebool\: \pair{\Ele{}{}}{\pair{\V{x}}{\V{y}}{}}{}                        & \F{leHost}\: \V{x}\: \V{y} 
\end{array} \\
\\
\evalH : \PI{\V{ty}}{\Ty} 
           \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}} \To
           \Val{\V{ty}} \\
\evalH\: \V{ty}\: \V{term} = \F{cataI}\: \Ty\: 
                                 \toIDesc{\HExprD}\: 
                                 \Val{}\: 
                                 \evalOne\: 
                                 \V{ty}\: 
                                 \V{term}
}\]
\note{pwm: $\evalOne$ rather illustrates my point about the 2 types of
  IFunc floating around, $\idescop{\_}{}{}$ is defined using the 1st
  notion, and used it here as if it were the 2nd. pierre: in this
  case, it was a typo. But you're right, there is two notions and we
  probably want to clarify where these things live.}

@@

However, so far, we are only able to define and manipulate
\emph{closed} terms. By abstracting over $\Val{}$, it is possible to
build and manipulate \emph{open} terms, that is terms with symbolic
variables. On the model of $\Val{}$, we define $\Var{}{}$:

@@

Consequently, replacing $\Val{\V{ty}}$ by
$(\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}})$ in Figure~\ref{fig:hexpr-full} turns
the language of closed term into a language of opened terms with
constants. For readability, we will abbreviate $\LAM{\V{ty}}
\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}}$ into $\SUM{\Val{}}{\Var{}{}}$ \note{Is
  it polite and comprehensible to ask that?} This defines a new
indexed description, called $\HExprVarD{\V{dom}}$.

@@

\[
\begin{array}{@{}ll}
\substH  : & \PITEL{\V{\dom}}{\EnumU} \\
           & \PITEL{\V{\gammaNat}}{\GammaSpi{\V{dom}}{\Enat}} 
             \PITEL{\V{\gammaBool}}{\GammaSpi{\V{dom}}{\Ebool}} \\
           & \begin{array}{@{}ll}
             \PI{\V{\sigma}}{& \PITEL{\V{dom}}{\EnumU} \\
                             & \PITEL{\V{\gammaNat}}
                                     {\GammaSpi{\V{dom}}{\Enat}} \\
                             & \PITEL{\V{\gammaBool}}
                                     {\GammaSpi{\V{dom}}{\Ebool}} \\
                             & \PI{\V{ty}}{\Ty} (\SUM{\Val{\V{ty}}}{\Var{\V{dom}}{\V{ty}}}) \To \\
                             & \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}}}
             \end{array}\\
          & \PI{\V{ty}}{\Ty}
            \IMu{\Ty}{\toIDesc{\HExprVarD{\V{dom}}}}{\V{ty}} \To \\
          & \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}}
\end{array}
\]

@@

Abstracting away the book-keeping introduced by the context, this
definition looks familiar. Indeed, it is extremely similar to a
monadic \bind. This is not surprising as we are defining a first-order
syntax with variables: our data-type enjoys more structure than what
we are given. In particular, this is reminiscent to a free monad,
where $\Eval{}$ is the \return\ introducing variables. The substitution
$\V{\sigma}$ is implemented from $\discharge$, by picking the domain and
context corresponding to the type:

@@

Instead of implementing $\substH$ in this special case, we can
implement it in a generic setting, a \emph{free indexed-monad construction}
that we will see in the next section.

@@

In this section, we develop more structure for our indexed
data-types. In Section~\ref{sec:desc-free-monad}, we have built a free
monad operation for simple descriptions. Building on this experience,
we are going to present its equivalent in the indexed world. The
process the same. Namely, given an indexed functor, we derive the
indexed functor coding its free monad: \note{pwm: Whoa there. Maybe we
  should say something about IMonads in general before we get to this
  point?}
